::: {.d-flex .gap-2 .justify-content-start .mb-4}

<a href="https://github.com/DanieleT25/ASML-Stock-Price-Analysis" target="_blank" class="btn btn-outline-secondary btn-sm" role="button">
  <i class="bi bi-github"></i> GitHub Repository
</a>

<a href="https://danielet25.github.io/ASML-Stock-Price-Analysis/presentation.html" target="_blank" class="btn btn-outline-primary btn-sm" role="button">
  <i class="bi bi-easel"></i> View Slides
</a>

<a href="https://danielet25.github.io/ASML-Stock-Price-Analysis/ASML-Stock-Price-Analysis.pdf" target="_blank" class="btn btn-outline-danger btn-sm" role="button">
  <i class="bi bi-file-earmark-pdf"></i> Download PDF
</a>

:::

# Preface {.unnumbered}

```{r}
#| label: setup
#| include: false

source(here::here("_setup.R"))
asml <- readRDS(here::here("data", "asml_data.rds"))
```

This project aims to conduct an in-depth time series analysis of the stock prices for **ASML Holding NV (ASML)**, a global leader in the manufacturing of lithography systems for the semiconductor industry. The primary objective is to explore the dynamics of historical prices, volatility, and underlying trends of the stock.

## Data Source

We retrieve historical daily data using the `quantmod` R package, focusing on the **Adjusted Closing Price** to account for dividends and stock splits.

* **Ticker**: ASML
* **Source**: [Yahoo Finance](https://finance.yahoo.com/quote/ASML/)
* **Start Date**: 2000-01-01
* **End Date**: 2025-12-31

## Data Loading

The dataset structure is shown below:

```{r}
#| label: show-table-data

datatable(
  tail(asml, 100),
  rownames = FALSE,
  options = list(
    pageLength = 10,
    order = list(list(0, 'desc'))
  )
)
```
